### 2024-10-01 teaching/math240f24/draft.tex
lecture october 1

\vskip \medskipamount \hbox to\hsize {\relax \unhcopy \strutbox \cont
{{\tenssbx I. Foundations \kern 3pt}}{2}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 1.\kern .5em Set theory }{3}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 2.\kern .5em Propositional logic}{10}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 3.\kern .5em Predicate logic}{17}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 4.\kern .5em Proofs}{21}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 5.\kern .5em Functions}{28}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
%\smallskip
%\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}

@@

The set of all infinite binary strings is in bijection with the set of all real numbers in the interval
$[0,1)$. If we're being extra pedantic, we need to forbid an infinite trailing string of all $1$s,
since, e.g., $0.0111\ldots = 0.1000\ldots$ after carrying the $1$s (akin to the fact
that, e.g., $0.0999\ldots = 0.1$). After dealing with this detail, one will have shown that
$\bigl| [0,1]\bigr| \ne |\NN|$.
\bye

@@

\datestamp{10}{} A {\it relation} on a set $X$ is a subset $R\subseteq X\times X$. If $(a,b)\in R$
we write $aRb$ and say ``$a$ is related to $b$.'' Here are some examples.
\medskip
\item{i)} The set $L = \bigl\{(a,b)\in \RR\times\RR : a<b\bigr\}$ is a relation. For instance, we can
write $(2,\pi)\in L$, or $2L\pi$, or $2<\pi$ (which is the more common notation for this relation).
\smallskip
\item{ii)} The set $E = \bigl\{ (a,b) \in \RR\times \RR : a = b\bigr\}$ is a relation (it is the ``equals''
relation).
\smallskip
\item{iii)} On $\ZZ$, the set
$$R = \bigl\{(-1, 4), (8, -3), (0, 0), (0,1)\bigr\}$$
is a relation.
\smallskip
\item{iv)} For any set $A$, a function $f:A\to A$ is by definition a subset of $A\times A$, and hence
is an example of a relation.
\smallskip
\item{v)} Let $H$ be the set of all humans, define $M\subseteq H\times H$ by setting
$(h_1, h_2)$ if and only if $h_1$ is married to $h_2$. For instance,
$$(\hbox{Michelle Obama}, \hbox{Barack Obama})\in M.$$
On the other hand, it is unfortunately the case that
$$(\hbox{Marcel Goh}, \hbox{Taylor Swift})\notin M.$$
\medskip

@@

Take for instance the example $F$ above of all formulas under the relation $\equiv$. The equivalence class
of the formula $p\vee\neg p$ is the set of all formulas whose truth table contains only $1$s, that is,
the set of all tautologies.
And for the example $R$ above, the
equivalence class of the point $(1,3)$ is the set
$$\bigl[ (1,3)\bigr] = \bigl\{ (x,y)\in \RR^2 : \sqrt\{10\} = \sqrt{x^2 + y^2}\bigr\};$$
that is, the circle of radius $\sqrt{10}$ centred about the origin.

@@

\proof Let $x\in A$. Since $R$ is reflexive, $x\sim x$, so we have $x\in [x]$. This proves (i).

@@

On the other hand, suppose that $x,y\in A$ satisfy $x\sim y$.
By (i) again, we have $[x] = [y]$, so $[x] \cap [y] = [x] \ne \emptyset$,
where we know that $[x]\ne \emptyset$ because it contains at least $x$.\slug

\noindent
\datestamp{11}{} \hskip\parindent Let $A$ be a set. A {\it partition} of $A$ is a set $P$ of subsets
of $A$ (i.e., $P\subseteq 2^A$), such that
\medskip
\item{i)} for all $x\in A$ there exists $S\in P$ such that $x\in S$;
\smallskip
\item{ii)} for all $S_1, S_2\in P$ with $S_1\ne S_2$, the intersection $S_1\cap S_2$ is empty; and
\smallskip
\item{iii)} $\emptyset\notin P$.
\medskip
Here are some examples.
\medskip
\item{i)} Let $E$ be the set of all even integers and $F$ the set of all odd integers. Then $\{E,F\}$
is a partition of $\ZZ$.
\smallskip
\item{ii)} The set $\bigl\{ (-\infty,0), \{0\}, (0,\infty)\bigr\}$ is a partition of $\RR$.
\medskip
If $\sim$ is an equivalence relation on a set $A$, then we can define the {\it quotient}
of $A$ by $\sim$ as the set of all equivalence classes of $A$ under $\sim$. We denote this set by
$$A/\sim = \bigl\{ [x] : x\in A\bigr\}.$$
We use the previous proposition to prove that equivalence relations give rise to partitions.

\proclaim Proposition \advthm. Let $A$ be a set and $\sim$ an equivalence relation on $A$.
Then $A/\sim$ is a partition of $A$.

>>>>

\vskip \medskipamount \hbox to\hsize {\relax \unhcopy \strutbox \cont
{{\tenssbx I. Foundations \kern 3pt}}{2}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 1.\kern .5em Set theory }{3}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 2.\kern .5em Propositional logic}{10}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 3.\kern .5em Predicate logic}{17}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 4.\kern .5em Proofs}{21}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 5.\kern .5em Functions}{28}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}

@@

The set of all infinite binary strings is in bijection with the set of all real numbers in the interval
$[0,1)$. If we're being extra pedantic, we need to forbid an infinite trailing string of all $1$s,
since, e.g., $0.0111\ldots = 0.1000\ldots$ after carrying the $1$s (akin to the fact
that, e.g., $0.0999\ldots = 0.1$). After dealing with this detail, one will have shown that
$\bigl| [0,1]\bigr| \ne |\NN|$.

@@

\datestamp{01}{x} A {\it relation} on a set $X$ is a subset $R\subseteq X\times X$. If $(a,b)\in R$
we write $aRb$ and say ``$a$ is related to $b$.'' Here are some examples.
\medskip
\item{i)} The set $L = \bigl\{(a,b)\in \RR\times\RR : a<b\bigr\}$ is a relation. For instance, we can
write $(2,\pi)\in L$, or $2L\pi$, or $2<\pi$ (which is the more common notation for this relation).
\smallskip
\item{ii)} The set $E = \bigl\{ (a,b) \in \RR\times \RR : a = b\bigr\}$ is a relation (it is the ``equals''
relation).
\smallskip
\item{iii)} On $\ZZ$, the set
$$R = \bigl\{(-1, 4), (8, -3), (0, 0), (0,1)\bigr\}$$
is a relation.
\smallskip
\item{iv)} For any set $A$, a function $f:A\to A$ is by definition a subset of $A\times A$, and hence
is an example of a relation.
\smallskip
\item{v)} Let $H$ be the set of all humans, define $M\subseteq H\times H$ by setting
$(h_1, h_2)$ if and only if $h_1$ is married to $h_2$. For instance,
$$(\hbox{Michelle Obama}, \hbox{Barack Obama})\in M.$$
On the other hand, it is unfortunately the case that
$$(\hbox{Marcel Goh}, \hbox{Taylor Swift})\notin M.$$
\medskip

@@

Take for instance the example $F$ above of all formulas under the relation $\equiv$. The equivalence class
of the formula $p\vee\neg p$ is the set of all formulas whose truth table contains only $1$s, that is,
the set of all tautologies.
And for the example $R$ above, the
equivalence class of the point $(1,3)$ is the set
$$\bigl[ (1,3)\bigr] = \bigl\{ (x,y)\in \RR^2 : \sqrt{10}= \sqrt{x^2 + y^2}\bigr\};$$
that is, the circle of radius $\sqrt{10}$ centred about the origin.

@@

\proof Let $x\in A$. Since $R$ is reflexive, $x\sim x$, so we have $x\in [x]$. This proves part (i).

@@

On the other hand, suppose that $x,y\in A$ satisfy $x\sim y$.
By (ii), we have $[x] = [y]$, so $[x] \cap [y] = [x] \ne \emptyset$,
where we know that $[x]\ne \emptyset$ because it contains at least $x$ (again, using (i)).\slug

Let $A$ be a set. A {\it partition} of $A$ is a set $P$ of subsets
of $A$ (i.e., $P\subseteq 2^A$), such that
\medskip
\item{i)} for all $x\in A$ there exists $S\in P$ such that $x\in S$;
\smallskip
\item{ii)} for all $S_1, S_2\in P$ with $S_1\ne S_2$, the intersection $S_1\cap S_2$ is empty; and
\smallskip
\item{iii)} $\emptyset\notin P$.
\medskip
Here are some examples.
\medskip
\item{i)} Let $E$ be the set of all even integers and $F$ the set of all odd integers. Then $\{E,F\}$
is a partition of $\ZZ$.
\smallskip
\item{ii)} The set $\bigl\{ (-\infty,0), \{0\}, (0,\infty)\bigr\}$ is a partition of $\RR$.
\medskip
\bye
If $\sim$ is an equivalence relation on a set $A$, then we can define the {\it quotient}
of $A$ by $\sim$ as the set of all equivalence classes of $A$ under $\sim$. We denote this set by
$$A/{\sim} = \bigl\{ [x] : x\in A\bigr\}.$$
We use the previous proposition to prove that equivalence relations give rise to partitions.

\proclaim Proposition \advthm. Let $A$ be a set and $\sim$ an equivalence relation on $A$.
Then $A/{\sim}$ is a partition of $A$.
